The regularity of ODEs and thimble integrals \\ with respect to Borel summation

### 2024-03-18 exp-resurgence.tex
Drop old text from intro to ODE results

Second of all, we want to explain why it is useful to work in the position domain (the Borel plane): on the one hand integral equations are more regular than differential equations. On the other hand, a thimble integral in the frequency domain can be recast as the Laplace transform of a function in the spatial domain. 
\subsubsection{Why does Borel summation work for solutions of level $1$ ODEs?}\label{sec:why_borel_ODE}
%\textcolor{DarkCyan}{[Let us try to move some technical detail from here to the section where the result is proved. For example:]}
%
Consider a linear level-$1$ differential operator $\mathcal{P}$ of the form described in~Section~\ref{borel-reg:explanatory-power}. This operator will always have an irregular singularity at $z = \infty$.

\color{Goldenrod}
[These details are right above in Section~\ref{borel-reg:explanatory-power}!]
\begin{equation}%\label{eqn:standard ODE}
\mathcal{P} = P(\partial_z) + \frac{1}{z} Q(\partial_z) + \frac{1}{z^2}\big[ R_0(z^{-1}) + R_1(z^{-1})\,\partial_z + \ldots + R_{d-1}(z^{-1})\,\partial_z^{d-1} \big],
\end{equation}
where $P$ is a monic degree-$d$ polynomial, $Q$ is a degree-$(d-1)$ polynomial, and $R_j(z^{-1})$ are holomorphic in some disk $|z| > A$ around $z = \infty$, for every $j=0,\ldots, d-1$.\footnote{Looking at the existence theorem in \cite[Theorem 4]{reg-sing-volterra}, we could apply this reasoning on the analytic side for more general equations, but this particular case makes it easier to talk about the formal side as well.} Furthermore we assume $P$ has simple zeros $P(-\alpha_j)=0$, $j=1,...,d$, $P'(-\alpha_j)\neq 0$ for $j=1,...,d$ and $Q(-\alpha_j)\neq 0$. Our equation has an irregular singularity at $\infty$ (see \cite[Definition 3.3.2]{diverg-resurg--ii}).
\color{black}
\begin{verify}
    Turn the ODE into a system: assume 
    \begin{align*}
P(\lambda) & = P_0 + P_1 \lambda + \ldots + P_d \lambda^d \\
Q(\lambda) & = Q_0 + Q_1 \lambda + \ldots + Q_{d-1} \lambda^{d-1}
\end{align*}
then the ODE $\mathcal{P}\psi=0$ is equivalent to the system 
    \begin{equation}
        \frac{d}{dz}Y= B(z) Y 
    \end{equation}
    \[
    B(z)=\begin{pmatrix}
            0 & 1 & 0 & \ldots & 0 \\
            0 & 0 & 1 & \ldots & 0 \\
            \vdots & &\ddots & & 1\\
            b_{1}(z) & & b_j(z) & \ldots & b_{d-1}(z)
        \end{pmatrix}, \quad Y=\begin{pmatrix}
            \psi \\
            \psi ' \\
            \vdots \\
            \psi^{(d-1)}
        \end{pmatrix}
    \]

@@

and $-w^{-2}B(1/w)$ has a simple pole at $w=0$ (i.e. $z=\infty$). Then the formal system of solutions is of the form 
\[\psi_j(w)=e^{-\alpha_j/w} w^{\tau_j} \tilde{f}_j(w)\]
hence according to \cite[Definition 3.3.2]{diverg-resurg--ii}, $w=0$ is irregular. 
\end{verify}

\color{gray}
[This is now a corollary; moved to historical.] Under these assumptions, the space of formal solutions of $\mathcal{P}\Phi = 0$ has a basis $\series{\Psi}_1,...,\series{\Psi}_d$ of the form~\cite{int-irreg}\cite[Proposition~2.2.7, p.~111]{EcalleIII}
\begin{equation}\label{formal solution}
\series{\Psi}_j(z)=e^{-\alpha_j z}z^{-\tau_j}\series{F}_j(z)\in e^{-\alpha_j z } z^{-\tau_j}\,\C \llbracket z^{-1} \rrbracket,
\end{equation}
where $\tau_j$ is the non-zero complex number $Q(-\alpha_j)/P'(-\alpha_j)$. Since the roots $-\alpha_1, \ldots, -\alpha_n$ are distinct, this form determines the basis up to scaling, giving the space of formal solutions a distinguished frame. \textcolor{orange}{Add that the formal series is $1$-Gevrey.} Classically it was proved that the formal solutions of \eqref{formal solution} are Borel-Laplace summable and their Borel-Laplace sums satisfy the original differential equation (see for example \cite{ramis1991series,malgrange92,malgrange1995sommation,diverg-resurg--ii}\textcolor{magenta}{[Braaksma,check]}). Therefore, the distinguished frame of formal solutions $\series{\Psi}_1,...,\series{\Psi}_d$ becomes a distinguished frame of analytic solutions $\hat{\Psi}_1,...,\hat{\Psi}_d$.

However, can we find a distinguished basis for solutions in a purely analytic way? This should be possible because of our regularity assumptions about the coefficients of $\mathcal{P}$. In fact, the Main Asymptotic Existence Theorem (M.A.E.T.) guarantees the existence of an analytic frame of solutions asymptotic to a formal frame (see \cite[Theorem 3.1]{diverg-resurg-iii}), but it is not a constructive method (see \cite[Chapter 14]{balser}). \color{black} \color{Goldenrod}[Let us try to move most of this detail to the proof section.] Conversely, we can explicitly build a distinguished basis of analytic solutions: first going to the position domain, we can consider $d$ integral operators of the form 
\begin{equation}
\hat{P}_j:=P(-\zeta)+\partial_{\zeta,\alpha_j}^{-1}\circ Q(-\zeta)+\partial_{\zeta,\alpha_j}^{-2}\circ\sum_{r=0}^{d-1}R_r(\partial_{\zeta,\alpha_j}^{-1})\circ (-\zeta)^r
\end{equation}
labeled by the roots of $P$. This type of operator is the one we studied in \cite{reg-sing-volterra}, hence we know there exists a unique solution $\psi_j$ of $\hat{P}_j\psi_j=0$, such that $\psi_j$ blows-up as $\zeta_j^{\tau_j-1}$ at $\zeta_j=0$ and it has a controlled behavior on an open domain $\Omega_j$ which is star-shaped at $\alpha_j$ and avoids the other critical points (see the definition in Section \ref{borel_reg-ODE}). In addition, $\psi_j$ can be Laplace transformed, giving a analytic function $\Psi_j:=\laplace_{\zeta_j}^{\theta}\psi_j$. 
At this point, we may ask what is the relationship between the two frames of analytic solutions, namely the Borel sum $\hat{\Psi}_j$ of the formal solutions $\series{\Psi}_j$ and the Laplace transform of $\psi_j$. Our main result says they are actually proportional, hence proving Borel summation works \textcolor{orange}{[add reminder of what it means for Borel summation to ``work'']}.\color{black}
\textcolor{Cyan}{Using the dictionary for the Laplace transform associated with some point $\zeta = \alpha$ in the frequency domain, (link section), we can turn $\mathcal{P}$ into an integral operator $\hat{\mathcal{P}}_\alpha$. If we choose $-\alpha$ to be a root of $P$, then $\hat{\mathcal{P}}_\alpha$ will have a regular singularity at $\zeta = \alpha$, and we understand a lot of details about its specific form~\cite{reg-sing-volterra}. We show in~\cite{reg-sing-volterra} that $\hat{\mathcal{P}}_\alpha \phi = 0$ has a distinguished solution $\psi_\alpha$ which is characterized by having a power-law singularity with a known exponent $\tau_\alpha - 1$ at $\zeta = \alpha$ and having exponentially bounded growth at infinity.}

We know from experience that we can find holomorphic solutions of the equation $\mathcal{P}\Phi = 0$ by taking the Borel sums of formal trans-monomial solutions. However, this only works when there's a coincidence between exponential factor $e^{-\alpha z}$ in the formal solution, the base point $\zeta = \alpha$ we use for Borel summation, and the characteristic equation $P(-\alpha) = 0$. Why?

@@

\textcolor{red}{[Change or actually delete]}We summarize the main steps of the proof of Theorem~\ref{thm:borel_reg_ODE} and Corollary~\ref{cor:borel_reg_ODE} (the rigorous proofs are given in Section~\ref{borel_reg-ODE}). On the right-hand side of the diagram, we have the formal solution $\series{\Psi }_\alpha\in e^{-\alpha z}z^{-\tau_\alpha}\C\llbracket z^{-1}\rrbracket$ and its Borel transform $\series{\psi}_\alpha$. %Then, the theory of $1$-summability allows us to draw the ``square'' with corners $\hat{\psi}_j$ and its Laplace transform $\hat{\Psi}_j$.

@@

Conversely, on the left-hand side, by our existence and uniqueness result \cite[Theorem 4]{reg-sing-volterra} we have an analytic function $\psi_\alpha$ with a distinguished behaviour near $\zeta=\alpha$, which solves the integral equation (IE) obtained as the inverse Laplace transform of the (ODE) (see \cite[Lemma 2]{reg-sing-volterra}). The key step is then comparing $\psi_\alpha$ and $\tilde{\psi}_\alpha$. First, by the properties of the Borel transform, it is easy to show that $\series{\psi}_\alpha$ is a formal solution of the integral equation (IE). Then, from the explicit formula for $\tilde{\Psi}_\alpha$, we can show that if $\tilde{\psi}_\alpha$ grows as $\psi_\alpha$ near $\zeta=\alpha$. Therefore, by our uniqueness result \cite[Theorem 4]{reg-sing-volterra}, we conclude that $\tilde{\psi}_j$ is indeed summable and ${\psi}_\alpha$ is the analytic continuation of $\tilde{\psi}_\alpha$.

\subsubsection{Why does Borel summation work for thimble integrals?}
%\textcolor{DarkCyan}{Let $X$ be an $N$-dimensional complex manifold, and $f\colon X\to\C$ a proper holomorphic map with non-degenerate critical points.}
Consider a thimble integral of the form described in Section~\ref{borel-reg:explanatory-power}. We will define our setting in the language of translation surfaces, to make clear the connection with our geometric picture of the Laplace transform. The codomain of $f$ has a natural structure of translation surface---a Riemann surface carrying a holomorphic 1-form $d\zeta$, and $\zeta$ is a translation coordinate. \textcolor{RoyalBlue}{[The important thing about $\C$, where $f$ takes values, is that it is a translation surface---\ldots]} For each critical point $a$, let $\Gamma_\alpha^\theta$ be the ray going to infinity with an angle $\theta$ from the critical value $\alpha$, and let $\zeta_{\alpha}$ be the translation coordinate around $\Gamma_\alpha^\theta$ which vanishes at $\alpha$. These are well-defined as long as $\Gamma_\alpha^\theta$ misses the critical values of $f$. Then, the preimage $f^{-1}(\Gamma_\alpha^\theta)$ is a bunch of disjoint curves (see Figure \ref{fig:thimble_vs_rays}), as long as $\Gamma_\alpha^\theta$ misses the other critical values of $f$. The Lefschetz thimble $\Lambda_a^\theta$ can be regarded as the component of $f^{-1}(\Gamma_\alpha^\theta)$ that goes through $a$.  \textcolor{orange}{We could orient so that shifting $\Lambda_a$ to its left would make its projection run clockwise around $\Gamma_\alpha$. However, this only seems to be possible when $X$ is 1-dimensional.}

>>>>

Second of all, we want to explain why it is useful to work in the position domain (the Borel plane): on the one hand integral equations are more regular than differential equations. On the other hand, a thimble integral in the frequency domain can be recast as the Laplace transform of a function in the spatial domain. 
\subsubsection{Why does Borel summation work for solutions of level $1$ ODEs?}\label{sec:why_borel_ODE}
%\textcolor{DarkCyan}{[Let us try to move some technical detail from here to the section where the result is proved. For example:]}
%
Consider a linear level-$1$ differential operator $\mathcal{P}$ of the form described in~Section~\ref{borel-reg:explanatory-power}. This operator will always have an irregular singularity at $z = \infty$.
\begin{verify}
\par
Turn the ODE into a system: assume 
\begin{align*}
P(\lambda) & = P_0 + P_1 \lambda + \ldots + P_d \lambda^d \\
Q(\lambda) & = Q_0 + Q_1 \lambda + \ldots + Q_{d-1} \lambda^{d-1}
\end{align*}
then the ODE $\mathcal{P}\psi=0$ is equivalent to the system 
    \begin{equation}
        \frac{d}{dz}Y= B(z) Y 
    \end{equation}
    \[
    B(z)=\begin{pmatrix}
            0 & 1 & 0 & \ldots & 0 \\
            0 & 0 & 1 & \ldots & 0 \\
            \vdots & &\ddots & & 1\\
            b_{1}(z) & & b_j(z) & \ldots & b_{d-1}(z)
        \end{pmatrix}, \quad Y=\begin{pmatrix}
            \psi \\
            \psi ' \\
            \vdots \\
            \psi^{(d-1)}
        \end{pmatrix}
    \]

@@

and $-w^{-2}B(1/w)$ has a simple pole at $w=0$ (i.e. $z=\infty$). Then the formal system of solutions is of the form 
\[\psi_j(w)=e^{-\alpha_j/w} w^{\tau_j} \tilde{f}_j(w)\]
hence according to \cite[Definition 3.3.2]{diverg-resurg--ii}, $w=0$ is irregular.
\par
\end{verify}
We know from experience that we can find holomorphic solutions of the equation $\mathcal{P}\Phi = 0$ by taking the Borel sums of formal trans-monomial solutions. However, this only works when there's a coincidence between exponential factor $e^{-\alpha z}$ in the formal solution, the base point $\zeta = \alpha$ we use for Borel summation, and the characteristic equation $P(-\alpha) = 0$. Why?

@@

\textcolor{red}{[Change or actually delete]}
\color{Violet}We summarize the main steps of the proof of Theorem~\ref{thm:borel_reg_ODE} and Corollary~\ref{cor:borel_reg_ODE} (the rigorous proofs are given in Section~\ref{borel_reg-ODE}). On the right-hand side of the diagram, we have the formal solution $\series{\Psi }_\alpha\in e^{-\alpha z}z^{-\tau_\alpha}\C\llbracket z^{-1}\rrbracket$ and its Borel transform $\series{\psi}_\alpha$. %Then, the theory of $1$-summability allows us to draw the ``square'' with corners $\hat{\psi}_j$ and its Laplace transform $\hat{\Psi}_j$.

@@

Conversely, on the left-hand side, by our existence and uniqueness result \cite[Theorem 4]{reg-sing-volterra} we have an analytic function $\psi_\alpha$ with a distinguished behaviour near $\zeta=\alpha$, which solves the integral equation (IE) obtained as the inverse Laplace transform of the (ODE) (see \cite[Lemma 2]{reg-sing-volterra}). The key step is then comparing $\psi_\alpha$ and $\tilde{\psi}_\alpha$. First, by the properties of the Borel transform, it is easy to show that $\series{\psi}_\alpha$ is a formal solution of the integral equation (IE). Then, from the explicit formula for $\tilde{\Psi}_\alpha$, we can show that if $\tilde{\psi}_\alpha$ grows as $\psi_\alpha$ near $\zeta=\alpha$. Therefore, by our uniqueness result \cite[Theorem 4]{reg-sing-volterra}, we conclude that $\tilde{\psi}_j$ is indeed summable and ${\psi}_\alpha$ is the analytic continuation of $\tilde{\psi}_\alpha$.
\color{black}
\subsubsection{Why does Borel summation work for thimble integrals?}
%\textcolor{DarkCyan}{Let $X$ be an $N$-dimensional complex manifold, and $f\colon X\to\C$ a proper holomorphic map with non-degenerate critical points.}
Consider a thimble integral of the form described in Section~\ref{borel-reg:explanatory-power}. We will define our setting in the language of translation surfaces, to make clear the connection with our geometric picture of the Laplace transform. The codomain of $f$ has a natural structure of translation surface---a Riemann surface carrying a holomorphic 1-form $d\zeta$, and $\zeta$ is a translation coordinate. \textcolor{RoyalBlue}{[The important thing about $\C$, where $f$ takes values, is that it is a translation surface---\ldots]} For each critical point $a$, let $\Gamma_\alpha^\theta$ be the ray going to infinity with an angle $\theta$ from the critical value $\alpha$, and let $\zeta_{\alpha}$ be the translation coordinate around $\Gamma_\alpha^\theta$ which vanishes at $\alpha$. These are well-defined as long as $\Gamma_\alpha^\theta$ misses the critical values of $f$. Then, the preimage $f^{-1}(\Gamma_\alpha^\theta)$ is a bunch of disjoint curves (see Figure \ref{fig:thimble_vs_rays}), as long as $\Gamma_\alpha^\theta$ misses the other critical values of $f$. The Lefschetz thimble $\Lambda_a^\theta$ can be regarded as the component of $f^{-1}(\Gamma_\alpha^\theta)$ that goes through $a$.  \textcolor{orange}{We could orient so that shifting $\Lambda_a$ to its left would make its projection run clockwise around $\Gamma_\alpha$. However, this only seems to be possible when $X$ is 1-dimensional.}
